### 2021-11-12 documents/sections/introduction.tex
updated readme
tables for introduction
many changes and additions to introduction

\input{stuff/cariban_tree}

>>>>

\subsection{The Cariban language family}

%todo write something here about the family in general
%maybe a nice little map-cum-tree, eh?
%\input{floats/cariban_tree}

\subsection{Verbs with conservative first person forms}

\input{floats/hixintro}
\input{floats/triintro}
Cariban languages feature relatively rich verbal morphology, both pre- and suffixes, inflecting for person, number, tense, aspect, and evidentiality, combined with a range of valency-modifying derivational markers.
Person paradigms for four \hixka \gl{s_a_} verbs are given in \cref{tab:hixintro}.%
\footnote{The presence of a \gl{1+2} person value implies that of a \gl{1+3} value.
This is usually expressed with a free pronoun combined with third person morphology in Cariban languages, so it is not represented as a distinct value in the paradigms we show.}%
\footnote{In \cref{tab:hixintro} and the remainder of this paper, we omit any TAM suffixes found in the original forms found in the literature, since a) our focus lies on the prefixes and stems, and b) full paradigms containing the same TAM suffix are rarely found.}%
\footnote{
We use standard IPA symbols in our transcription of Cariban languages, with the exception of coronal rhotics, which we simply represent with \ort{r}, rather than \ort{ɽ} for \wayana or \ort{ɾ̠} for \maqui etc.
In languages with strong morphophonological processes and/or subphonemic orthography we show the original transcription in an additional surface line when presented in an interlinearized glossed example.
We follow \textcite{gildea2018reconstructing} in using \ort{ə} for the proto-vowel reconstructed by \textcite{meira2005southern}, although it was likely more back \parencite{gildea2010story}.}
It is evident that the verb \qu{to be} diverges from otherwise identically inflected verbs by having a different first person marker.
That prefix is \obj{w-}, while other \gl{s_a_} verbs like \qu{to fall}, have \obj{k(ɨ)-}.
Another similar example comes from \trio, where the verb \qu{to go} has a first-person prefix \obj{wɨ-} while other \gl{s_a_} verbs have a prefix with phonologically conditioned allomorphs \envr{\obj{t-}}{\obj{ə}} and \envr{\obj{s-}}{\obj{e}} \pcref{tab:triintro}.
From a synchronic, language-internal perspective, one could call these prefixes and the verbs on which they appear \textsc{irregular}, contrasting with regular prefixes on regular \gl{s_a_} verbs, \obj{k(ɨ)-} and \obj{t-}/\obj{s-}.
However, there is no widely accepted definition of irregularity \parencite{stolz2012introduction}, and many of the more strict ones \parencite[][e.g.]{haspelmath2010understanding} require the pattern to occur at a single place in the grammar.
However, both languages have a few other verbs patterning the same way; for example, the first-person form of \trio \qu{to be} is \obj{w-ei-}.
Whether or not they fit one's understanding of irregular, these verbs can be analyzed as belonging to a small inflection (sub-)class. %
%\footnote{That analysis is found in synchronic descriptions of other Cariban languages discussed in more detail later, see \textcites[149]{alves2017arara}[4]{meira2003bakairi}.}
To answer the question of how these inflectional patterns came into being, one needs to turn to diachrony, starting with the person marking system in \PC.

%
%\begin{inlinelist}
%	\item special copula prefix \parencite[188]{hixkaryanaderby1985}
%	\item inflectional subclass \parencites[149]{alves2017arara}[4]{meira2003bakairi}
%	\item \dbqu{exceptional cases} \parencite[293]{triomeira1999}
%	\item phonologically conditioned allomorphs \parencites[139]{meira2006syntactic}{meira2003primeras}[189]{hixkaryanaderby1985}
%	\item not discussed \parencites{waiwaihawkins1998}{ikpengpacheco2001}{alves2013verbo}{pacheco2003intransitivos}
%\end{inlinelist}.s

%meira: non-detransitivized verbs
%alves: non-detransitivized verbs (as a consequence, o-initial)

%\subsection{Cariban verbal person marking and marker extensions}
%\label{sec:background}
%The archaic first person markers investigated in this paper are the result of incomplete person marker extensions, i.e., modifications to the ancestral prefix system that did not affect the whole lexicon.
%To provide readers with the necessary background, we first introduce said ancestral system \pcref{sec:pc_person}, then the peculiarities of the Cariban split-\gl{s} system \pcref{sec:split}, and finally the concept of person marker extensions \pcref{sec:extensions_intro}.

\subsubsection{\PC verbal person marking}
\label{sec:pc_person}
\PC is reconstructed by \textcite{gildea1998} as using a person paradigm called \setone in its independent verb forms, shown in \cref{tab:pcpers}.%
The choice of person marker in transitive verbs can be characterized as being conditioned by a basic person hierarchy \fbox{\gl{1}/\gl{2} > \gl{3}}.\footnote{Glossing abbreviations: }%\glossingAbbrevsComma.}
The locuphoric markers had two forms, an \gl{a}-oriented one for direct (\gl{sap}>\gl{3}) scenarios and a \gl{p}-oriented one for inverse (\gl{3}>\gl{sap}) scenarios.
There was a single aliophoric marker \rc{n(i)-}, which only surfaced in nonlocal (\gl{3}>\gl{3}) scenarios, without morphologically expressed distinctions between different third person referents.
Local scenarios were expressed in a non-transparent manner, by using the \gl{1+2} prefix \rc{k-} in both cases.%

\input{floats/pcpers}

Formally identical or etymologically related markers occured in intransitive verbs, which showed a split-\gl{s} system \pcref{tab:pcintrans}.
That is, \gl{s_a_} verbs took similar markers as the \gl{a}-oriented ones in transitive verbs, with the exception of first person (\gl{1}>\gl{3} \rc{t(i)-} vs \gl{1}\gl{s_a_} \rc{w-}), as well as the absence of \rc{i} after all \gl{s_a_} prefixes.
On the other hand, \gl{s_p_} verbs took markers fully identical to the \gl{p}-oriented ones.
The third person marker in \gl{s_p_} verbs was identical to the one in \gl{3}>\gl{3} scenarios (\rc{n(i)-)}.

From this diachronic perspective, it becomes evident that both \hixka \obj{w-} and \trio \obj{wɨ-} are reflexes of the \PC \gl{1}\gl{s_a_} prefix \rc{w-}.
Notably, it is reconstructed as the only \gl{1}\gl{s_a_} prefix, a situation still found in other modern languages like \maqui \pcref{tab:makintro}.
This means that \hixka \qu{to be} and \trio \qu{to go} are divergent because they preserve the old prefix, while other verbs have taken on new prefixes.
From a language-specific point of view, these verbs are thus \textsc{conservative}, contrasting with innovative verbs.
The prefixes may be considered \textsc{relics}, old and restricted to specific lexically conditioned contexts, contrasting with the innovative prefixes found elsewhere.
One can draw a parallel to the few English nouns like \obj{ox}, which preserve the old plural suffix \obj{-en}.
It, too, was once more widespread, being the normal plural suffix of the weak inflection, compare German \obj{Ochse-n}, \obj{Mensch-en}, \obj{Hase-n}.
Having established the conservative nature of these prefixes, the obvious question is where the new prefixes came from.

\input{floats/makintro}

\subsubsection{Person marker extensions in intransitive verbs}
\label{sec:extensions_intro}
The original \PC split-\gl{s} system has been subject to change in many languages, in some even to the point of total loss.
Besides modifications like losing the \gl{s_a_} class marker \rc{w-} \pcref{sec:split} or losing certain person values altogether, these changes have largely been due to person prefixes being extended to new verbs.%
\footnote{
In principle, one could also imagine a scenario in which verbs of one of the classes are gradually replaced with innovative verbs from the other class, as suggested for \panare by \textcite[225]{meira2000split}.
However, the verbs Meira sampled came from the \emp{A}-section of \posscite{mattei1994diccionario} dictionary \perscommpar{Spike Gildea}, and \obj{at-}\slash{}\obj{at͡ʃ-}\slash{}\obj{as-} is a frequent reflex of \detrz in \panare, meaning that many \gl{s_a_} verbs are \obj{a}-initial, thus explaining the 80:40 \gl{s_a_}:\gl{s_p_} ratio in Meira's sample.
Thus, the hypothesis that \panare has largely replaced \gl{s_p_} with \gl{s_a_} verbs remains to be thoroughly tested.
For the moment, the most likely scenario leading to the eventual loss of the split-\gl{s} system involves markers being extended to new verbs.}
There have been many such person marker extensions in Cariban languages, and some are still ongoing.
This is shown by \textcite{gildea1998}, using the three Parukotoan languages as an example.
We have reproduced his tables as a tree diagram in \cref{fig:par_ext}, with adapted transcription and the added \kaxui alternative \gl{1}\gl{s_p_} marker ∅/\obj{j-} \perscommpar{Spike Gildea}.
Apart from segmental changes to individual morphemes, the following restructuring innovations happened in the \setone paradigm in Parukotoan:

\begin{enumerate}
	\item \PPar \begin{enumerate}
		\item \gl{1}\gl{s_a_} \rc{w-} to \gl{1}>\gl{3}
		\item \gl{1+2} \rc{k-} to \gl{1}\gl{s_p_} (completed in \PWai, ongoing in \kaxui)
		\item \gl{1+2} \rc{kɨt-} to \gl{1+2}\gl{s_p_} (completed in \PWai, ongoing in \kaxui)
	\end{enumerate}
	\item \PWai \begin{enumerate}
		\item \gl{1}\gl{s_p_} \rc{k-} to \gl{1}\gl{s_a_}
		\item innovative \rc{owɨro j-} \qu{\gl{1}\gl{pro} \gl{lk}} for \gl{1}\gl{p}
	\end{enumerate}
	\item \waiwai \begin{enumerate}
		\item \gl{2}\gl{s_a_} \obj{m-} to \gl{2}\gl{s_p_}\end{enumerate}
\end{enumerate}

\begin{figure}[hbt]
	\centering
	\setlength{\tabcolsep}{2pt}
	\begin{tikzpicture}[
		every node/.append style={align=center},
		level distance=80pt,
		]
		\Tree[.{\PC{}\pcone}
		[.{\PPar{}\ppone}
		{\kaxui{}\kaxone}
		[.{\PWai{}\pwone}
		{\hixka{}\hixone}
		{\waiwai{}\waione}
		]
		]
		]
	\end{tikzpicture}
	\caption{Person marking extensions in Parukotoan, after \textcite[94]{gildea1998}}
	\label{fig:par_ext}
\end{figure}

\hixka has preserved split-\gl{s} only in the second person prefixes, while \kaxui still shows variation in the first person and \gl{1+2} prefixes.
\waiwai, on the other hand, has lost the system entirely, which notably happened via distinct innovations at three different diachronic stages.
In this case, the loss of the inflectional classes also entailed the loss of the other morphological traces of the split-\gl{s} system.
That is, the \gl{s_a_} class marker \rc{w-} \pcref{sec:split} was lost in \waiwai, as evidenced by the contrast between the \waiwai and \kaxui deverbal forms in \exref{wailoss.wai-158} and \exref{wailoss.kax-138}.
Similarly, the \gl{2}\gl{s_p_} prefix \obj{a-} was extended to imperatives of (former) \gl{s_a_} verbs \exref[wailoss.wai-112]{wailoss.hix-43}, but only C-initial ones \parencite[62]{waiwaihawkins1998}.

\pex<wailoss>
\begin{multicols}{2}
	\a<wai-158> \waiwai \parencite[][98]{waiwaihawkins1998}\\
	\begingl
	\gla k-eɸɨrka-t͡ʃhe//
	\glb \gl{1+2}-fall-\gl{advz}.after//
	\glft \qu{after our fall}//
	\endgl
	\a<kax-138> \kaxui \parencite[][49]{schuring2018kaxuyana}\\
	\begingl
	\gla ku-w-ehurka-tpɨrɨ//
	\glb \gl{1+2}-\gl{s_a_}-fall-\gl{nmlz}.\gl{pst}.\gl{pert}//
	\glft \qu{our fall}//
	\endgl
	\a<wai-112> \waiwai \parencite[][177]{waiwaihawkins1998}\\
	\begingl
	\gla a-mo-ko//
	\glb \gl{2}-come-\gl{imp}//
	\glft \qu{come!}//
	\endgl
	\a<hix-43> \hixka \parencite[][191]{hixkaryanaderby1985}\\
	\begingl
	\gla m-omokɨ-no//
	\glb \gl{2}\gl{s_a_}-come-\gl{imm}//
	\glft \qu{You have come.}//
	\endgl
\end{multicols}
\xe

%Such extensions, often only affecting one person value at a time, and the eventual loss of split-\gl{s} are not entirely unexpected, given that the system lacks any semantic basis.

While different cases of loss of split-\gl{s} by person marking extensions are discussed by \textcite[91--96]{gildea1998}, this paper focuses on a so far neglected aspect of these extensions.
As a prerequisite for our story, we argue that they are take place via lexical diffusion, characterized as a type of extension by \textcite[106--115]{harris1995historical}; this hypothesis is supported by three facts.
First, the variation in first person and \gl{1+2} prefixes described above for \kaxui is not completely free.
Rather, some verbs only allow for example first person \obj{k-}, but not \obj{j-}, while others can occur with both, which is the expected pattern in a lexical diffusion scenario.
In addition, this is speaker-dependent \perscommpar{Spike Gildea}, which is what one would expect from a change in progress.
%todo maybe put in GOOD kaxui data
Second, while there is no detailed diachronic scenario for the switch of \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} in the Tiriyoan languages \pcref{sec:taranoan}, \textcite[111--112]{meira1998proto} argues that it must have happened gradually rather than instantaneously, and entailed both markers spreading at the same time.
Whether this gradual switch was along ordered lines or not, lexical diffusion must have played a role.
Our final argument in favor of the lexical diffusion scenario brings us back to the \hixka and \trio verbs in \cref{tab:triintro,tab:hixintro}:
We know from synchronic \hixka that some \gl{s_a_} verbs take \obj{w-} rather than \obj{k-}.
We interpret this as the extension of \gl{1}\gl{s_p_} \rc{k-} to \gl{1}\gl{s_a_} which began in \PWai stopping short of these verbs, and not spreading through the entire verbal lexicon.

We investigated 19 person marker extensions affecting intransitive verbs, and identified 6 which left a group of verbs untouched, like \PWai \gl{1}\gl{s_p_} \rc{k-} to \gl{1}\gl{s_a_}.
These have left between 1 and 7 conservatively inflected verbs in 9 Cariban languages, marked with bold print in \cref{fig:caribantree}.
Interestingly, these extensions always feature innovative first person markers on \gl{s_a_} verbs.
While extensions occurred with other person values as well, they never affected \gl{s_a_} verbs, only \gl{s_p_} ones, and they always affected the entire lexicon, at least based on the attested data.
Illustrative examples for extensions affecting the entire lexicon are shown in \cref{tab:completed}: the extension of \gl{1+2}\gl{s_a_} \obj{s(ɨ)-} (< \rc{kɨt-}) in \apalai \pcref{tab:apalai}, of \gl{2}\gl{s_a_} \obj{m(ɨ)-} in \panare \pcref{tab:panare},\footnote{The presence of the third person marker \obj{n-} for \gl{1+2} is due to the wholesale loss of that inflectional value.} or the extension of the entire \gl{s_a_} set in \waimiri \pcref{tab:waimiri}.%, as well as the extension of \rc{k-} to \gl{1}\gl{s_p_} and of \rc{tɨt-} to \gl{1+2}\gl{s_p_} in \PWai \pcref{fig:par_ext}.

The different behavior of \gl{s_a_} and \gl{s_p_} verbs with regards to incomplete extensions points to the split-\gl{s} system playing some role.
Thus, we will introduce that system before discussing the incomplete extensions in more detail.
This will also give readers an idea how it is possible for the \gl{s_a_}--\gl{s_p_} distinction to be lost only for a single person, or for \gl{s_p_} verbs to take on \gl{s_a_} markers with apparent semantic impunity.

\input{floats/completed}

\subsubsection{The \PC split-\gl{s} system}
\label{sec:split}
As seen in \cref{sec:pc_person}, the \PC distinction between \gl{s_a_} and \gl{s_p_} verbs was made by two inflectional classes for intransitive verbs within the \setone system, but this was not the only morphological criterion:
%\begin{inlinelist}
%	\item presence vs absence of the \gl{s_a_} marker \rc{w-}
%	\item absence vs presence of the second person prefix \rc{ə(j)-} in imperatives
%	\item presence vs absence of a derivational detransitivizing prefix
%\end{inlinelist}.
%
Many languages show an \gl{s_a_} class marker in deverbalized forms, which can be reconstructed to \PC as \rc{w-}.\footnote{See \textcite[227]{meira2000split}, who identifies reflexes of this morpheme as having \dbqu{no purpose other than being \qu{class markers}, without any obvious semantic or functional load}.}
With \gl{s_a_} verbs, \rc{w-} occurred immediately between the possessive prefixes and the verb stem, while \gl{s_p_} verbs took the bare prefixes.
Reflexes of \rc{w-} in languages from different branches are illustrated in \cref{tab:participles,tab:nominalizations} for participles and nominalizations.%
%
\input{floats/participles}%
%
\input{floats/nominalizations}%
%
The distinction between \gl{s_a_} and \gl{s_p_} was also borne out in imperatives.
Here, \gl{s_p_} verbs took the \gl{p}-oriented second person prefix \rc{ə(j)-}, while \gl{s_a_} verbs were unprefixed (both suffixed with \rc{-kə}).
This is illustrated with reflexes in various modern languages in \cref{tab:imperatives}.
Both the \gl{s_a_} marker \rc{w-} in participles and nominalizations and the prefixes in imperativees have been lost in some languages, but are reconstructible to \PC.

\input{floats/imperatives}

%\subsubsection{The detransitivizing prefix}
There is one further property uniting most \gl{s_a_} verbs, which is not based on inflectional morphology.
As mentioned in \cref{sec:pc_person}, mismatches between the semantics of intransitive verbs and their \gl{a}- or \gl{p}-oriented inflectional morphology are common.
However, the Cariban split-\gl{s} system goes further than all other known such systems, in that its division of the verbal lexicon does not follow any discernible semantic criteria whatsoever.
\textcite{meira2000split} takes a sizable corpus of intransitive verbs from \trio, \kalina, \apalai, and \wayana, and categorizes them by applying different criteria commonly encountered in split-\gl{s} systems.
He shows that neither (non\-)activities, % \parencite[210]{meira2000split}, 
(non-)agency, % \citeyearpar[212]{meira2000split}, 
(in-)animacy, % \citeyearpar[213]{meira2000split},
nor Aktionsart % \citeyearpar[215]{meira2000split} 
satisfactorily predict the class membership of intransitive verbs.

Rather, the reason for a verb to take the \gl{a}- or \gl{p}-oriented prefix is (at least diachronically) a morphological one.
\textcite[217--221]{meira2000split} demonstrates that those intransitive verbs which (etymologically) have a detransitivizing prefix are treated as \gl{s_a_} verbs, while essentially all others are \gl{s_p_} verbs:
\begin{quotebox}{\parencite[201]{meira2000split}}
	Almost all verbs in the \gl{s_a_} class are detransitivized forms of transitive verbs, either synchronically (with still exisiting transitive sources) or diachronically (with reconstructible but no longer existing transitive sources)
\end{quotebox}
\textcite[221--223]{meira2000split} also argues that the detransitivizing prefixes are indeed deriving \gl{s_a_} verbs, rather than being inflectional in nature:
\begin{inlinelist}
	\item there are a few underived \gl{s_a_} verbs, with no detransitivizing prefix
	\item \gl{s_a_} verbs can develop irregular semantics compared to their transitive counterparts
	\item it is unpredictable whether the \gl{a} or \gl{p} argument of the underlying transitive verb becomes the \gl{s} of the derived \gl{s_a_} verb
	\item some originally derived \gl{s_a_} verbs have lost their transitive counterparts
	\item \dbqu{basic} concepts are expressed as derivations of more complex concepts, like \qu{to dance (\gl{s_a_})} from \qu{to dance with (\gl{tr})}
\end{inlinelist}.
He also notes that this leads to an inflectional split not based in meaning, but rather morphology:

\begin{quotebox}{\parencite[226]{meira2000split}}
	Apparently, the morphological behavior of the \gl{s_a_} verb class is an accidental consequence of the fact that detransitivization, as far back as we can reconstruct, entails all the morphology described […] as typical of \gl{s_a_} verbs. The alignment of person-marking prefixes appears not to be driven by any semantic forces in the language; it is as though they were being dragged by the evolution of the reflexive marker.
\end{quotebox}

As for the form of this marker, \textcite[505--512]{meira2010origin} reconstruct two distinct  prefixes for \PC: reciprocal \rc{əte-} and reflexive \rc{e-}, although they have since merged into a single morpheme, apparently in all languages.
Modern reflexes of \detrz show a range of meanings, which can all be characterized as \dbqu{detransitive}; this range is illustrated with \trio examples in \exref{triodetrz}.

\ex<triodetrz> \trio \parencites[218--219]{meira2000split}[128, 256]{triomeira1999}\\
\begin{tabular}[t]{@{}llll@{}}
	\\
	\obj{nonta}  & → & \obj{e-nonta}, & \qu{abandon each other}\\
	\qu{abandon} & & \obj{əi-nonta} &  (reciprocal) \\
	\\
	\obj{suka} & → & \obj{e-suka}, & \qu{wash self}\\
	\qu{wash} & & \obj{əi-suka} & (reflexive)\\
	\\
	\obj{pahka} & → & \obj{e-pahka} & \qu{break (\gl{intr})}\\
	\qu{break (\gl{tr})} & & & (anticausative)\\
	\\
	\obj{puunəpɨ} & → & \obj{əh-puunəpɨ}, & \qu{think, meditate}\\
	\qu{think about} & & \obj{əi-puunəpɨ} & (antipassive)\\
\end{tabular}
\xe
%
The morphological variation featured in \qu{to abandon each other} and \qu{to wash self} is due to the mentioned collapse between the two \PC prefixes:
\obj{e-} is a reflex of the reflexive prefix \rc{e-}, while the form \obj{əi-} originates in reciprocal \rc{əte-}.
However, both can occur with either meaning -- at least for these two verbs.

As mentioned, there are a few \gl{s_a_} verbs that are not derived from transitive verbs, and do not have a reflex of \detrz, either.
\textcite[221]{meira2000split} characterizes these as very old and exceptional, and counts between 5 and 7 verbs, depending on the language.
\textcite{gildea2007greenberg} identify 7 such underived \gl{s_a_} verbs as reconstructible to \PC, shown in \cref{tab:underived2007}.
Many of these verbs will be discussed in \cref{sec:verbs}.

\input{floats/underived2007.tex}

As in many other split-\gl{s} systems found in languages of the world, the intransitive verbal lexicon was divided into two classes, with verbs inherently being members of the \gl{s_a_} or the \gl{s_p_} group.
That is, while in transitive verbs, the choice of person marker had a crucial semantic contribution, it was predictable in intransitive verbs.
This is illustrated with modern \kalina data in \exref[kar1]{kar2}.

\pex<kar1> \kalina
\a<kar-60>
\begingl
\gla mi-kupi-ja//
\glb \gl{2}>\gl{3}-bathe-\gl{prs}//
\glft \qu{You bathe him/her.} \parencite[][160]{hoff1968carib}//
\endgl
\a<kar-62>
\begingl
\gla a-kupi-ja//
\glb \gl{3}>\gl{2}-bathe-\gl{prs}//
\glft \qu{S/he bathes you.} \parencite[][63]{yamada2011evidentiality}//
\endgl
\xe
%
In \exref{kar1}, the choice between the second person \gl{a}- and \gl{p}-oriented markers \obj{mi-} and \obj{a-} depends on the scenario:
The transitive verb \obj{kupi} \qu{to bathe} takes \obj{mi-} in \gl{2}>\gl{3} scenarios \exref{kar1.kar-60}, but \obj{a-} in \gl{3}>\gl{2} scenarios \exref{kar1.kar-62}.
While intransitive verbs show the same (or very similar) person markers, they contribute no semantic difference here \exref{kar2}.

\pex<kar2> \kalina
\a<kar-59>
\begingl
\gla sipi tɨnka-rɨ m-ekema-non hen//
\glb net pull-\gl{nmlz} \gl{2}-be.afraid-\gl{prs}.\gl{uncert} eh?//
\glft \qu{You're afraid to pull up the net, aren't you? } \parencite[][253]{courtz2008carib}//
\endgl
\a<kar-61>
\begingl
\gla aj-awoi-ja//
\glb \gl{2}-get.up-\gl{prs}//
\glft \qu{You are getting up.} \parencite[][167]{hoff1968carib}//
\endgl
\xe
%
Rather, \obj{ekema} \qu{to be afraid} takes an \gl{a}-oriented marker, since it is an \gl{s_a_} verb \exref{kar2.kar-59}, while the \gl{s_p_} verb \obj{awomɨ} \qu{to get up} takes a \gl{p}-oriented marker \exref{kar2.kar-61}.\footnote{The root \obj{awomɨ} \qu{to get up} is subject to syllable reduction and assimilation to the prefix-initial \obj{j}.}
What is more, the choice of prefix does not necessarily convey information about the semantics of the verb, since there are clear mismatches:
\qu{to be afraid} with an \dbqu{agentive} marker can hardly be considered a volitional act, while  \qu{to get up} with a \dbqu{patientive} marker must be considered volitional.
These mismatches are not isolated cases, as will be discussed in \cref{sec:split}.

This paper is dedicated to these conservatively inflected verbs, their origins, their similarities, and the reasons behind them.
It is structured as follows: \cref{sec:background} provides readers with the necessary background on Cariban verbal person marking, especially split-\gl{s} system, and introduces the mechanism by which the innovative prefixes spread to almost all \gl{s_a_} verbs.
The topic of \cref{sec:extensions} are the six individual innovations which introduced new first person prefixes, but left a small group of verbs unaffected.
Since these verbs show considerable etymological overlap between languages, they are reconstructed and discussed in more detail in \cref{sec:verbs}.
Finally, we search for factors motivating the resistance of these verbs, discuss the findings and put them in a general context of language change and morphology in \cref{sec:discussion}.
